Reproducible Research, Peer Assessment 1      
Coursera, October 2015      
Ken Gosier

Begin by setting echo=TRUE, and results as "asis"
globally for the document

```{r setoptions,echo=TRUE}
opts_chunk$set(echo = TRUE, results="asis")
```

## Loading and preprocessing the data

Here I unzip the file if necessary, and read in the
data through read.csv. Convert steps and interval to integer,
and the date column to type Date. Leave missing data in the
data frame for now.

```{r}
zipfile <- "activity.zip"
unzipfile <- "activity.csv"

if (!file.exists(unzipfile)) {
    unzip(zipfile, overwrite = TRUE)
}

data <- read.csv(unzipfile, header = TRUE,
    stringsAsFactors = FALSE)
data$steps <- as.integer(data$steps)
data$date <- as.Date(data$date, format="%Y-%m-%d")
data$interval <- as.integer(data$interval)
```

## What is mean total number of steps taken per day?

Here I aggregate the total number of steps taken each day.
NA's are ignored, effectively treating those intervals
as if no steps had been during those intervals, since they
will not contribute to the sum.

```{r}
aggbydate <- aggregate(data$steps, by = list(date = data$date),
    FUN = sum, na.rm = TRUE)
colnames(aggbydate) <- c("date", "total_steps")
```

Make a histogram. R's barplot command produced the wrong
output. It showed the total number of steps for each
day, rather than aggregating the frequency of steps across
the days, as hist does. I show 10 breaks in the histogram.

```{r}
hist(aggbydate$total_steps, xlab = "Total Steps per day",
    breaks = 10, main = "Histogram, Total Steps per day")
```

The mean of steps per day:

```{r}
mean(aggbydate$total_steps)
```

The median of steps per day:


```{r}
median(aggbydate$total_steps)

```

## What is the average daily activity pattern?

Take the average by interval across all days. At this point
we are still ignoring the NA's.

```{r}
aggbyinterval <- aggregate(data$steps, by = list(date = data$interval),
    FUN = mean, na.rm = TRUE)
colnames(aggbyinterval) <- c("interval", "average_steps")
```

Show a plot of average steps vs interval.

```{r}
plot(aggbyinterval$interval, aggbyinterval$average_steps, type = "l",
    xlab = "interval", ylab = "average_steps")
```

The index of the interval with the highest average steps.

```{r}
maxind <- which.max(aggbyinterval$average_steps)
maxind
```

The interval with the highest average steps.

```{r}
aggbyinterval$interval[maxind]

```

The number of steps in the max interval.

```{r}
aggbyinterval$average_steps[maxind]
```

## Imputing missing values

First calculating the total number of rows with missing values
in the dataset.

```{r}
length(which(is.na(data$steps)))
```

Replace missing values with the average value for that interval
across all the days. This is an over-simple way to fill in
the missing values.

It would be nice to do this replacement all in one statement
rather than a for loop, may look into lapply in the future.

```{r}
data.nona <- data
for (i in 1:nrow(aggbyinterval)) {
    thisint <- aggbyinterval$interval[i]
    thismean <- aggbyinterval$average_steps[i]
    thissubs <- which(is.na(data.nona$steps) & data.nona$interval == thisint)
    data.nona$steps[thissubs] <- thismean
}
```

Aggregate data.nona by total steps per day, and make a histogram.

```{r}
aggbydate.nona <- aggregate(data.nona$steps, by = list(date = data.nona$date),
    FUN = sum)
colnames(aggbydate.nona) <- c("date", "total_steps")
hist(aggbydate.nona$total_steps, xlab = "Total Steps per day",
    breaks = 10, main = "Histogram, NA's replaced by interval mean")
```

Computing the mean and median of steps per day with the NA's replaced.

```{r}
mean(aggbydate.nona$total_steps)
```

```{r}
median(aggbydate.nona$total_steps)
```

Both have increased from their previous values. This makes sense -- with
the NA's remove from the sums, those intervals were essentially treated
as if they had 0 steps. We have replaced these values with the average of
steps for the interval, so the mean and median must increase.

It can be seen the bin with 10,766 steps grows. There were several days with
NA listed for all the intervals. These all got moved from the left of the
histogram to the middle bin. In general the histogram moves a little to the
right -- any NA was previously effectively treated as 0, since it did
not contribute to the sum. Now there is a nonzero value in most cases
to add into the sum.

## Are there differences in activity patterns between weekdays and weekends?

Create a factor which tells whether each day was a weekday or weekend.

```{r}
wdayvec <- character(nrow(data.nona))
wdayvec[1:length(wdayvec)] <- "weekday"
data.weekdays <- weekdays(data.nona$date)
wdayvec[which(data.weekdays == "Saturday" | data.weekdays == "Sunday")] = "weekend"
data.nona$weekend_indic <- as.factor(wdayvec)
```

Average data by weekend indicator and interval, and make a panel plot
showing average steps per interval through the day for weekends and
weekdays.

```{r}
aggbyinterval.weekday <- aggregate(data.nona$steps,
    by = list(interval = data.nona$interval, weekend_indic = data.nona$weekend_indic),
    FUN = mean, na.rm = TRUE)
colnames(aggbyinterval.weekday) <- c("interval", "weekend_indic", "average_steps")
library(lattice)
xyplot(average_steps ~ interval | factor(weekend_indic),
    data = aggbyinterval.weekday,
    type = "l",
    xlab = "interval",
    ylab = "Number of steps",
    layout = c(1,2))
```

Just as an observation, the weekday days show more activity in the hours before
work starts at 9am, around lunchtime, around a mid-afternoon snack at 3pm, and
in the hours immediately after 5pm. The weekend steps are more spread out.
